Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota \\ {\large Senior Comprehensive Exercise: Carleton College}

### 2010-06-08 NewPaper.tex
Added the real discussion.

Our model performed surprisingly well, particularly given the complex and
sparse nature of the data analyzed. We were successfully able to incorporate
false detection, spatial data correlation, and the dependence of detection
probabilities on time since snowfall events into a model and still obtain both
accurate and precise results. While others have had success incorporating one
of these processes into statistical models \citep{Magoun2007, Royle2006,
Stanley2005}, we are not aware of any studies that have incorporated all three.
[Talk more about specifics when we get them.] Having the ability to model such
complex processes will make data collection in similar presence/absence surveys
both easier and less costly.

Interestingly, in our simulations with spatially correlated data, neither of
our two spatially explicit models performed particularly better than the
simpler spatially inexplicit model. While confidence intervals were sometimes
slightly smaller for the conditional autoregressive (CAR) models that took into
account the spatial correlation, the differences were small, and due to the
significantly longer computing times necessitated by the added complexity of
the spatial models, the simpler model may be a better choice. This is contrary
to what has been found in other studies that attempting to use CAR models with
similar binary data \citep{Wintle2006}.  However, the level of correlation in
their data may have been higher.

In addition, because funds for aerial surveys such as the ones used to collect
data in this study are limited, we explored different survey designs to
determine which aspects are most important in yielding precise and accurate
results. In particular, we examined numbers of observers, snow events, days
surveyed following snow events, and sites. [Make some sort of grand sweeping
conclusions about which are most important?] No single parameter appears most
important. However, we recommend a minimum sampling effort of [what?] to ensure
that the standard error for psi falls below [what?]. It is notable that if
there is more data available, the model will generally take less time to
converge.

These statistical methods could be used on data for other species, as well.
[John said we should talk about this. Which species?]

@@

\parindent -0.15in

Gorman, T., Erb, J., McMillan, B.\ \& Martin, D.\ (2006) Space Use and
Sociality of River Otters ({\em Lontra Canadensis}) in Minnesota. {\em Journal
of Mammalogy}, {\bf 87}, 740--747.

\parindent -0.15in
Mackenzie, D.\ \& Royle, J.\ (2005) Designing Occupancy Studies: General Advice
and Allocating Survey Effort, {\em Journal of Applied Ecology}, {\bf 42},
1105-1114.

>>>>

Our model performed surprisingly well, particularly given the complex and
sparse nature of the data analyzed. We were successfully able to incorporate
false detection, spatial data correlation, and the dependence of detection
probabilities on time since snowfall events into a model and still obtain both
accurate and precise estimates for the occupancy rate. While others have had
success incorporating one of these processes into occupancy models
\citep{Magoun2007, Royle2006, Stanley2005}, we are not aware of any studies
that have incorporated all three. [Talk more about specifics when we get them.]
Having the ability to model such complex processes will make data collection in
similar presence/absence surveys both easier and less costly.

Interestingly, in our simulations with spatially correlated data, neither of
our two spatially explicit models performed particularly better than the
simpler spatially inexplicit model. While confidence intervals were sometimes
slightly smaller for the conditional autoregressive (CAR) models that took the
spatial correlation into account, the differences were small, and due to the
significantly longer computing times necessitated by the added complexity of
the spatial models, the simpler model may be a better choice. This is contrary
to what has been found in other studies that attempted to use CAR models with
similar binary data \citep{Wintle2006}. However, the level of correlation in
their data may have been higher.

In addition, because funds for aerial surveys are limited, we explored
different survey designs to determine which aspects are most important in
yielding precise and accurate results. In particular, we examined numbers of
observers, snow events, days surveyed following snow events, and sites. No
single parameter appears most important. However, we recommend a minimum
sampling effort of [what?] to ensure that the standard error for psi falls
below [what?]. \citet{Mackenzie2005} found that when detection probabilities
exceed 0.5, as they do in the case of our data, they recommended a minimum of
three surveys. Because we separate the track-laying and detection processes
into two separate parameters ($p$ and $\theta$, respectively), more than three
surveys are necessary in studies like this one.

These statistical methods could be used on investigations of other species, as
well. Similar aerial snow-track surveys have been used in population studies
for other riparian species such as ptarmigan, snowshoe hares, and red foxes
\citep{St-Georges1995}, and non-riparian species such as caribou
\citep{Courtois2003} and wolverines \citep{Magoun2007}. Although there may not
be a natural river corridor to fly along for non-riparian species, transects on
which to fly could be drawn ad hoc. The would be worthwhile, as there is great
potential for applications of complex occupancy models such as this one in
studies of population dynamics for use in wildlife conservation and management.

@@

\parindent -0.15in
Gorman, T., Erb, J., McMillan, B.\ \& Martin, D.\ (2006) Space Use and
Sociality of River Otters ({\em Lontra Canadensis}) in Minnesota. {\em Journal
of Mammalogy}, {\bf 87}, 740--747.
